Continuous Deductive Reasoning over Streamed RDF Data

### 2015-11-10 article.tex
significant changes to the discussion of (Windowed) Continuous Datalog to cover multiple streams, multiple windows, and pipelining programs.

\begin{definition}[Continuous Datalog Program]
\labeldef{continuous datalog: CDP}

We initially take a {\em Continuous Datalog program} (henceforth a
C-Datalog program) $\Pi^C = \langle R, F^0, S \rangle$ to be a tuple
which comprises a finite set of rules $R$, a finite set of static fact
instances $F^0$ (all of which have the instant $0$) and an in-order
stream of fact instances $S$.

The model theory for a such a C-Datalog program broadly follows that
for a TA-Datalog program; we again take the Herbrand base of all fact
instances $\hat{H}$. Disregarding the order of fact instances in $S$,
a fact instance $e$ is entailed by a program $\Pi^C$ iff each
interpretation satisfying every clause in $\Pi^C$ also satisfies $e$.
However, we take the {\em entailment of the C-Datalog program} $\Pi^C$
to be an in-order stream $E^S$, where the order in which entailed fact
instances appear in $E^S$ depends on the order in which the fact
instances in their justifications appear in $S$. Given that that the
instant of an entailed fact instance is the maximum of the instants in
the fact instances that form the justification for that entailed fact
instance (as in \refdef{continuous datalog: instance justification}),
$E^S$ will also be an in-order stream.

However, this definition of a C-Datalog program does not take into
account any notion of windowing over the stream $S$. We define a {\em
  Windowed Continuous Datalog program} $\Pi^{C, r} = \langle R, F^0, S,
r \rangle$ to be a tuple with components as defined above for a
C-Datalog program, but with the addition of $r \in \mathbb{N}$ for the
range of the window over $S$, which effectively specifies that the
fact instances from $S$ are to contribute to the entailments of $\Pi^{C,r}$
from their instant for a duration of $r$.

We consider the entailment of the WC-Datalog program $\Pi^{C,r}$ not
as a single stream of fact instances (as for C-Datalog), but in three
parts: a set of fact instances $E^0 = cons^T(\langle R, F^0\rangle)$
that is the entailment of the TA-Datalog program $\langle R,
F^0\rangle$; a stream $P$ of fact instances entailed by the rules $R$ and
the fact instances in $F^0$ and $S$, ordered by the instant at which they
became justified (i.e. their instants); and a stream $N$ of those same fact
instances, ordered by the instant at which they ceased to be justified:
$\langle R, F^0, S, r \rangle \models \langle E^0, P, N \rangle$. $P$
and $N$ can be considered to be the continuous reasoning equivalents of
the streams produced by the \emph{IStream} and\emph{DStream} operators
respectively of the \emph{CQL} continuous query language \citep{CQL},
which create streams based on the additions to and deletions from a
relation. Note that $P$ is an in-order stream whose members are a subset
of those in $E^S$ for an equivalent C-Datalog program, while $N$ may be
an out-of-order stream.
\end{definition}

@@

{\nobreak\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}
If two sliding windows over the same in-order stream $S$ cover
non-overlapping time periods, then the set of fact instances entailed
by ${\langle R, W^{S,x}_i \rangle}$ is necessarily disjoint from the
set of fact instances entailed by $\langle R, W^{S,y}_{i-z} \rangle$,
where ${z \geqslant x}$.
\[
\{ e \mid \langle R , W^{S,x}_i \rangle \models e \} \cap \{ e \mid \langle R , W^{S,y}_{i-z} \rangle \models e \} = \varnothing
\]
\end{axiom}}

@@

We can reduce a WC-Datalog program $\Pi^{C,r} = \langle R, F^0, S, r
\rangle$ at an instant $i \in \mathbb{T}$ to its TA-Datalog equivalent by
taking the window $W^{S,r}_i$ over stream $S$ of range $r \in \mathbb{N}$;
we refer to this as the {\em instantaneous state} of the WC-Datalog program,
written as $\Pi^{C,r}_i = \langle R, F^0 \cup W^{S,r}_i \rangle$. As
previously shown in \refdef{continuous datalog: ta datalog program}, this
may then be further reduced to the Datalog program $\Pi = \langle R, strip(F^0 \cup W^{S,r}_i) \rangle$.

The instantaneous entailment of a $\Pi^{C,r}_i$ may be divided into two
semantically distinct portions: the \emph{persistent} portion is the
entailment of ${\langle R, F^0 \rangle \models E^0}$, as in a WC-Datalog
program, which is guaranteed to be part of the entailment of every state
due to the monotonicity of positive Datalog; the \emph{transitive} portion
is the set $E^T_{i}$ of entailed fact instances whose justifications are at
least partially contained in $W^{S,r}_i$. {\nobreak We define $E^T_i$ as:
\[
E^T_i = \{ x \mid \Pi^{C,r}_i \models x \land J^T \cup W^{S,r}_i \neq \varnothing \}
\]
where $J^T$ is a justification for the entailment of x by
$\Pi^{C,r}_i$.} By this definition and Axiom~\ref{axiom:continuous
  datalog: entailment disjointness}, $E^0$ is guaranteed to be
disjoint from every $E^T_i$ for every state $\Pi^{C,r}_i$ of every
WC-Datalog program $\Pi^{C,r}$.
\end{definition}

@@

{\nobreak\begin{axiom}
\label{axiom:continuous datalog: entailment precedes negation}
Fact instances may only be negated by some part of their justification
leaving the sliding window $W^{S,r}_{i}$, this necessarily being after
they were entailed.  As such, a window of any range ${x \in
  \mathbb{N}}$ over the stream of entailments $P$ at any given instant
${i \in \mathbb{T}}$ is disjoint from a window of any range ${y \in
  \mathbb{N}}$ over the stream of entailment negations $N$ at any
instant before ${i - x}$.
\begin{equation*}
  W^{P,x}_{i} \cap W^{N,y}_{i-z} = \varnothing
\end{equation*}
where $z \geqslant x$.
\end{axiom}}

As noted above in \refdef{continuous datalog: CDP}, the stream of
negations $N$ of a WC-Datalog program $\Pi^{C,r}$ may be an
out-of-order stream in which fact instances appear on the stream in
the order in which they cease to have any complete justifications.

Thus far the translation from instantaneous sets of entailments to
streams of entailments has been stated axiomatically as the definition
of those streams.

However, it follows from these axioms that the instantaneous set
$E^T_{i}$ of instances entailed at any given instant $i$ by any
$\Pi^{C,r} = \langle R, F^0, S, r \rangle$ may be reconstructed from
windows of sufficient range over its entailed streams $P$ and $N$ at
$i$:
\[ E^T_{i} = W^{P,x}_{i} \setminus W^{N,y}_{i} \]
where ${y \geqslant x \geqslant r}$. The minimum range $x$ is such to achieve
the \emph{complete} set of entailment instances $E^T_{i}$ of the
$\Pi^{C,r}_{i}$. In addition, the minimum range $y$ is such to achieve the
\emph{sound} set $E^T_{i}$. As such, a \emph{valid window} that
contains only and exactly those entailed instances $E^T_{i}$ at every
instant $i$ may be trivially maintained from the streams $P$ and $N$
continuously entailed by some $\Pi^{C,r}$.

In conclusion, given a Continuous Datalog program $\Pi^C = \langle R,
F^0, S \rangle$, we can apply to it a window of range $r$ to yield a
Windowed Continuous Datalog program $\Pi^{C,r} = \langle R, F^0, S, r
\rangle$. We can evaluate this window at an instant $i \in \mathbb{T}$
to yield a Time-Annotated Datalog program $\Pi^T = \langle R, F^0 \cup
W^{S,r}_i \rangle$, from we which can then strip the time annotations
to yield a conventional Datalog program $\Pi = \langle R, F
\rangle$. The relationship between these *-Datalog programs and their
entailments is expressed in the commutative diagram in
\reffig{CD-TD-D}.

>>>>

\begin{definition}[(Windowed) Continuous Datalog Program]
\labeldef{continuous datalog: CDP}

We initially take a {\em Continuous Datalog program} (henceforth a
C-Datalog program) $\Pi^C = \langle R, F^0, S \rangle$ to be a tuple
which comprises a finite set of rules $R$, a finite set of persistent/static
fact instances $F^0$ (all of which have the instant $0$, meaning that they
cannot appear in any state of any window, by
\refdef{continuous datalog: window} and Axioms~\ref{axiom:continuous datalog: window range leq 0} and
\ref{axiom:continuous datalog: window start leq 0}) and an in-order stream
of fact instances $S$.  It should be noted that, in practice, data may be
drawn from more than one source, and so $S$ may be a union of
the mappings represented by $n$ in-order streams $S_0,\dots,S_n$ without altering
the semantics presented here.

The model theory for a such a C-Datalog program broadly follows that
for a TA-Datalog program; we again take the Herbrand base of all fact
instances $\hat{H}$. Disregarding the order of fact instances in $S$,
a fact instance $\langle e, t \rangle$ is entailed by a program $\Pi^C$
iff each interpretation satisfying every clause in $\Pi^C$ also
satisfies $\langle e, t \rangle$. However, we take the {\em entailment
of the C-Datalog program} $\Pi^C$ to be an in-order stream $E^S$ rather
than a set of fact instances. Given that the instant associated with an
entailed fact instance is the maximum of those associated with
any part of its justification (as in
\refdef{continuous datalog: instance justification}), and $E^S$ is an
in-order stream, $E^S$ expresses the times at which each entailed fact
instance was first justified by $\Pi^C$.

However, this definition of a C-Datalog program does not take into
account any notion of windowing over the stream $S$, and so any fact of
which an instance occurs in $S$ will be considered true from the instant
of its occurrence in $S$.  While this is technically acceptable in the
case of sets of rules without constraints, if a constraint exists and its
body becomes satisfied by some $\Pi^C$ at some instant $i$ then the
program will be unsatisfiable from that instant on, making $E^S$ empty
from then on.  As such, we define a {\em Windowed Continuous Datalog program}
$\Pi^{C, r} = \langle R, F^0, \{\langle S_0, r_0 \rangle,\dots,\langle S_n, r_n \rangle\} \rangle$
(where $r = max(r_0,\dots,r_n)$) to be a tuple with components $R$ and $F^0$ as defined above for a
C-Datalog program, but with each part of $S = S_0 \cup \dots \cup S_n$
associated with the range $r_x \in \mathbb{N}$ of a sliding window over
$S_x$, for $x \in [0,n]$. This specifies that the fact instances
from $S_x$ are to contribute to the entailments of $\Pi^{C,r}$ from the instant
they occur in $S_x$ for a duration of $r_x$ consecutive instants.

We consider the entailment of the WC-Datalog program $\Pi^{C,r}$ not
as a single stream of fact instances (as for C-Datalog), but a triple: a
set of fact instances $E^0 = cons^T(\langle R, F^0\rangle)$ that is the
entailment of the TA-Datalog program $\langle R, F^0\rangle$; a stream
$P$ of fact instances entailed by the rules $R$ and the fact instances in
$F^0$ and $S$, ordered by the instant at which they became justified (i.e.
their instants); and a stream $N$ of those same fact instances, ordered by
the instant at which they \emph{ceased} to be justified (i.e. some part of
their justification left a sliding window $W^{S_x,r_x}$). We express this
entailment $\Pi^{C,r} \models \langle E^0, P, N \rangle$.
It should be noted that $P$ and $N$ can be considered to be the continuous
reasoning equivalents of the streams produced by the \emph{IStream} and
\emph{DStream} operators, respectively, of the \emph{CQL} continuous query
language \citep{CQL}, which create streams based on the additions to and
deletions from a relation.

Note that $P$ is an in-order stream whose members are a subset
of those in $E^S$ for an equivalent C-Datalog program, up until the instant $i$
at which a contradiction occurs in $S$, while $N$ may be an out-of-order
stream. Furthermore, the members of $P$ from the instant $i+1$ are a subset of
those in $E^{S\prime}$, where $\langle R,F^0,S^\prime \rangle \models E^{S\prime}$
and $S^\prime$ is the in-order stream composed of the mappings in $S$ to instants
greater than $i$. In this way, a WC-Datalog program is able to \emph{recover}
from contradictions in $S$.

@@

{\nobreak\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}
If two states of windows over the same in-order stream $S$ cover
non-overlapping time periods, then the set of fact instances entailed
by ${\langle R, W^{S,x}_i \rangle}$ is necessarily disjoint from the
set of fact instances entailed by $\langle R, W^{S,y}_{i-z} \rangle$,
where ${z \geqslant x}$.
\[
\{ e \mid \langle R , W^{S,x}_i \rangle \models e \} \cap \{ e \mid \langle R , W^{S,y}_{i-z} \rangle \models e \} = \varnothing
\]
\end{axiom}}

@@

We can reduce a WC-Datalog program $\Pi^{C,r} = \langle R, F^0,
\{ \langle S_0, r_0 \rangle,\dots,\langle S_n, r_n \rangle \} \rangle$
at an instant $i \in \mathbb{T}$ to its TA-Datalog equivalent by
applying the rules $R$ to the union of the persistent facts $F^0$ and all
finite states $W^{S_x,r_x}_i$ of windows $W^{S_x,r_x}$ over streams $S_x$ of range
$r_x \in \mathbb{N}$, for all $x \in [0,n]$; we refer to this as the {\em instantaneous state} of
the WC-Datalog program, written
$\Pi^{C,r}_i = \langle R, F^0 \cup W^{S_0,r_0}_i \cup \dots \cup W^{S_n,r_n}_i \rangle$.
As previously shown in \refdef{continuous datalog: ta datalog program}, this may then be further
reduced to the Datalog program
$\Pi = \langle R, strip(F^0 \cup W^{S_0,r_0}_i \cup \dots \cup W^{S_n,r_n}_i) \rangle$.

The instantaneous entailment of a $\Pi^{C,r}_i$ may be divided into two
semantically distinct portions: the \emph{persistent} portion is the
entailment of ${\langle R, F^0 \rangle \models E^0}$, as in a WC-Datalog
program, which is guaranteed to be part of the entailment of every consistent
state due to the monotonicity of positive Datalog; the \emph{transient} portion
is the set $E^T_{i}$ of entailed fact instances whose justifications are at
least partially contained in $W^{S_0,r_0}_i \cup \dots \cup W^{S_n,r_n}_i$.
{\nobreak We define $E^T_i$ as:
\[
E^T_i = \{ x \mid \Pi^{C,r}_i \models x \land J^T \cap \left( W^{S_0,r_0}_i \cup \dots \cup W^{S_n,r_n}_i \right) \neq \varnothing \}
\]
where $J^T$ is a justification for the entailment of $x$ by
$\Pi^{C,r}_i$.} By this definition and Axiom~\ref{axiom:continuous
  datalog: entailment disjointness}, $E^0$ is guaranteed to be
disjoint from every $E^T_i$ for every state $\Pi^{C,r}_i$ of every
WC-Datalog program $\Pi^{C,r}$.

An \emph{inconsistent state} of a C-Datalog program is any instantaneous
state $\Pi^{C,r}_i$ that is unsatisfiable, i.e. that contains at least
one contradiction.  Such a state has no entailment.

@@

% It should be noted that a the existence of a contradiction in some
% C-Datalog program $\langle R, F^0, S \rangle$ does not imply that any
% \emph{state} of the WC-Datalog $\Pi^{C,r} = \langle R, F^0, S, r \rangle$ will be
% inconsistent. Where a state $\Pi^{C,r}_i$ of $\Pi^{C,r}$ \emph{is} inconsistent,
% the transient portion $E^T_i$ of its entailment is empty, as an unsatisfiable
% TA-Datalog program entails nothing.  In this case, therefore, \emph{all}
% fact instances transiently entailed by $\Pi^{C,r}_{i-1}$ occur in $N$ at the $i$,
% and no fact instances will occur in $P$ from $i$ until the first instant $j > i$ at
% which $\Pi^{C,r}_j$ is satisfiable.  In reference to the stream $E^S$ entailed by
% the C-Datalog program $\langle R, F^0, S \rangle$, therefore, $P$ is an in-order
% stream whose members \emph{that are mapped to instants between $i$ } are a subset of those in $E^S$
% to instants greater than $i$

{\nobreak\begin{axiom}
\label{axiom:continuous datalog: entailment precedes negation}
Fact instances may only be negated by some part of their justification
leaving the sliding window $W^{S,r}_{i}$, this necessarily being after
they were justified.  As such, a window of any range ${x \in
  \mathbb{N}}$ over the stream of entailments $P$ at any given instant
${i \in \mathbb{T}}$ is disjoint from a window of any range ${y \in
  \mathbb{N}}$ over the stream of entailment negations $N$ at any
instant before ${i - x}$.
\begin{equation*}
  W^{P,x}_{i} \cap W^{N,y}_{i-z} = \varnothing
\end{equation*}
where $z \geqslant x$.
\end{axiom}}

% As noted above in \refdef{continuous datalog: CDP}, the stream of
% negations $N$ of a WC-Datalog program $\Pi^{C,r}$ may be an
% out-of-order stream in which fact instances appear on the stream in
% the order in which they cease to have any complete justifications.

Thus far the translation from instantaneous sets of entailments to
streams of entailments has been stated axiomatically as the definition
of those streams.
However, it follows from these axioms that the instantaneous set
$E^T_{i}$ of instances entailed at any given instant $i$ by any single-window program
$\Pi^{C,r} = \langle R, F^0, \{ \langle S, r \rangle \} \rangle$ may be reconstructed from
windows of sufficient range {\nobreak over its entailed streams $P$ and $N$ at
$i$:
\[ E^T_{i} = W^{P,x}_{i} \setminus W^{N,y}_{i} \]
where ${y \geqslant x \geqslant r}$}. The minimum range $x$ is such to achieve
the \emph{complete} set of entailment instances $E^T_{i}$ of the
$\Pi^{C,r}_{i}$. In addition, the minimum range $y$ is such to achieve the
\emph{sound} set $E^T_{i}$.  It should be noted that the proof holds in the
case of $n$ windows of varying range, with $x \geqslant max(r_0,\dots,r_n)$.
We capture this derivation in the form of a \emph{valid window}, as per
\refdef{continuous datalog: valid window}.

\begin{definition}[Valid Window] \labeldef{continuous datalog: valid window}

A \emph{valid window} contains only and exactly those instances $E^T_{i}$ entailed by
the state $\Pi^{C,r}_i$ at every instant $i \in \mathbb{T}$.  Such a window may be
trivially maintained from the streams $P$ and $N$ continuously entailed
by some $\Pi^{C,r}$: adding all fact instances occurring in $P$ at $i$ to the state
of the window at $i-1$, then removing all fact instances occurring on $N$ at $i$ from
the resultant set of fact instances, produces the state of the window at $i$.  Such
a window obviously has no definite range but is taken over two streams (e.g. $P$ and
$N$), and so is denoted $W^{P \setminus N}$.

The decomposition of streams into more sparse streams without loss of
information allows us to define the state of a valid window
$W^{P \setminus N}$ as a union of the states of sliding windows over
each of the singleton streams $S_{\langle f, t \rangle} =
(\langle f, t \rangle \mapsto t^\prime)$, where
$S(\langle f, t \rangle) = t^\prime$.  The range of the sliding
window over each $S_{\langle f, t \rangle}$ is then the difference between
the instant at which $\langle f, t \rangle$ occurs in $P$ and that
at which it occurs in $N$:
\[
  W^{P \setminus N}_i = \{ \langle f, t \rangle \mid \langle f, t \rangle \in W^{S_{\langle f, t \rangle},N(\langle f, t \rangle) - P(\langle f, t \rangle)}_i\}
\]
Treated this way, the entailment of one WC-Datalog program, e.g. that in
\refeqn{continuous datalog: example WC-Datalog program},
may then be taken as a part or all of the data in a successive WC-Datalog
program, without modification of the existing semantics or loss of information
regarding the validity of the entailments.  However, in the interest
of being concise, such a successive WC-Datalog program may express valid
windows alongside any sliding windows. For example, 
a program composed of the set of rules $R^\prime \neq R$, the set of base facts
$F^{0\prime} \neq F^0$, a sliding window of range $x \geqslant r$ over the
in-order stream $S^\prime \neq S$, and the entailment of
$\Pi^{C,r} \models \langle E^0, P, N \rangle$, would be expressed as in \refeqn{continuous datalog: successive WC-Datalog program}.
\begin{gather}
  \labeleqn{continuous datalog: example WC-Datalog program}
    \Pi^{C,r} \langle R, F^0, \{W^{S,r}\} \rangle \models \langle E^0, P, N \rangle \\
  \labeleqn{continuous datalog: successive WC-Datalog program}
    \Pi^{C,x} = \langle R^\prime, F^{0\prime} \cup E^0, \{ W^{P \setminus N}, W^{S^\prime,x} \} \rangle \models \langle E^{0\prime}, P^\prime, N^\prime \rangle
\end{gather}

In conclusion, given a Continuous Datalog program $\Pi^C = \langle R,
F^0, S_0 \cup \dots \cup S_n \rangle$, we can apply to it a set of
windows of range $r_0,\dots,r_n$ to yield a Windowed Continuous Datalog
program $\Pi^{C,r} = \langle R, F^0, \{ W^{S_0, r_0}, \dots,
W^{S_n, r_n} \} \rangle$. We can evaluate these windows at
an instant $i \in \mathbb{T}$ to yield a Time-Annotated Datalog program
$\Pi^T = \langle R, F^0 \cup W^{S_0,r_0}_i \cup \dots \cup W^{S_n,r_n}_i
\rangle$, from we which can then strip the time annotations to yield a
conventional Datalog program $\Pi = \langle R, F \rangle$. The
relationship between these \ast-Datalog programs and their
entailments is expressed in the commutative diagram in
\reffig{CD-TD-D}.
